---
title: "MedicalCostPersonal"
author: "Beyza Özen"
date: "2023-01-02"
output: pdf_document
---

### Import libraries

```{r warning=FALSE}
library(dplyr)
library(tidyverse)
library(readr)
library(janitor)
library(ggplot2)
library(psych)
library(rafalib)
```

### Read Data

```{r warning=FALSE}
url <- "https://raw.githubusercontent.com/beyzaozen/MedicalCostPersonal/main/insurance.csv"
data <- read_csv(url)
```

## Exploratory Data Analysis

-   General Information The data contains 7 columns with 1338 lines. 4 of them are numerical and 3 are characters.

```{r}
glimpse(data)
```

There is no column without a value. So no need for cleaning or replacing na values.

```{r warning=FALSE}
data %>% select(everything()) %>%  # replace to your needs
  summarise_all(funs(sum(is.na(.))))
```

There is one row repeating in the dataframe, so I deleted the repeating row

```{r warning=FALSE}
data %>% get_dupes()

data<- data %>% distinct()

```

Check for number of unique observations for categorical values. For column "smoker", it can be factorized since it has yes and no as values. For sex and region, we need to use one-hot encoding for conveniene.

```{r}
sapply(data, is.character)

table(data$sex)

table(data$smoker)

table(data$region)

```

```{r}
data %>%
  tabyl(region, sex) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting() 

data %>%
  tabyl(region, smoker) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting() 

data %>%
  tabyl(smoker, sex) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting() 
```

```{r}
data %>%   
  select_if(is.numeric) %>% 
  gather(var,value) %>% 
  ggplot(aes(x = value))+
  geom_histogram()+
  facet_wrap(~var, scales = "free")
```

Inspecting data properties for anomalies before plotting.

```{r}
data %>% select_if(is.numeric) %>% summary
```

Changing smoker column to numeric by replacing yes and no with 1 and 0 values.

```{r}
#data$smoker[data$smoker == "yes"]<-1
#data$smoker[data$smoker == "no"]<-0
#data$smoker<-as.integer(data$smoker)
```

```{r}
#data$sex[data$sex == "female"]<-1
#data$sex[data$sex == "male"]<-0
#data$sex<-as.integer(data$sex)
#sapply(data, is.character)
```

The oval-shaped object on each scatterplot is a correlation ellipse. It provides a visualization of how strongly correlated the variables are. The dot at the center of the ellipse indicates the point of the mean value for the x axis variable and y axis variable. The correlation between the two variables is indicated by the shape of the ellipse; the more it is stretched, the stronger the correlation. An almost perfectly round oval, as with bmi and children, indicates a very weak correlation

```{r}
pairs.panels(data[c("age", "bmi", "children", "charges")])
```

```{r}
  data %>% ggplot() +
    geom_point(aes(age, charges, colour = sex)) +
    geom_smooth(aes(age, charges), method  = "lm")
```

The graph below show a clear distinction between smokers and nonsmokers

```{r}
  data %>% ggplot() +
    geom_point(aes(age, charges, colour = smoker)) +
    geom_smooth(aes(age, charges), method  = "lm")
```

```{r}
ggplot(data,aes(age, charges, colour = smoker)) +
     geom_point() +
     geom_smooth(data = data %>% filter(charges<15000), method  = "lm", color = "yellow")+
     geom_smooth(data = data %>% filter(charges>30000), method  = "lm", color = "green")+
     geom_smooth(data = data %>% filter(charges>15000 & charges<30000), method  = "lm", color = "blue")+
    geom_text(aes(x = 30, y = 15000, label = "charges<15000"), color = "yellow")+ 
  geom_text(aes(x = 30, y = 45000, label = "charges>30000"), color = "green")+ 
  geom_text(aes(x = 30, y = 30000, label = "charges>15000 & charges<30000"), color = "blue")
```

```{r}
data %>% 
  mutate(charges_cut = cut(data$charges, c(-Inf, 15000, 30000, Inf), labels=c("<15000", "15000<_<30000", ">30000"))) %>% 
  tabyl(charges_cut, smoker) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting() 

table(charges_cut = cut(data$charges, c(-Inf, 15000, 30000, Inf), labels=c("low", "mid", "high")))
```

-   group one, charges \< 15k, represents 73% (979/1338) of customers and 99.3% are non-smokers.
-   group two, charges \> 15k but \< 30k, represents 14% (196/1338)of customers among which 58.7% are smokers.
-   group three, charges \> 30k, represents 12% of customers among which 98.8% are smokers.

The graph below shows that, we can divide the data as ( bmi\<30 or not ) and (smoker or not)

```{r warning=FALSE}
  data %>% ggplot() +
    geom_point(aes(bmi, charges, colour = age, size = smoker)) 
    # + geom_smooth(aes(bmi, charges), method  = "lm")
```

```{r}
data %>% filter(bmi<=30  & smoker == "yes") %>%
{  ggplot(.) +
    geom_point(aes(bmi, charges, colour = age)) +
    labs(title=paste("bmi vs charges for smoker = ", .$smoker[1]))}

data %>% filter(bmi<=30  & smoker == "no") %>%
{  ggplot(.) +
    geom_point(aes(bmi, charges, colour = age)) +
    labs(title=paste("bmi vs charges for smoker = ", .$smoker[1]))}

data %>% filter(bmi>30  & smoker == "yes") %>%
{  ggplot(.) +
    geom_point(aes(bmi, charges, colour = age)) +
    labs(title=paste("bmi vs charges for smoker = ", .$smoker[1]))}

data %>% filter(bmi>30  & smoker == "no") %>%
  {ggplot(.) +
    geom_point(aes(bmi, charges, colour = age)) +
    labs(title=paste("bmi vs charges for smoker = ", .$smoker[1]))}
```

```{r}

pairs.panels(data %>% 
               filter(bmi>30  & smoker == "yes") %>%
               select(c("age", "bmi", "children", "charges")),
             main = "bmi>30 vs charges for smoker = yes ")

pairs.panels(data %>% 
               filter(bmi>30  & smoker == "no") %>%
               select(c("age", "bmi", "children", "charges")),
             main = "bmi>30 vs charges for smoker = no ")

pairs.panels(data %>% 
               filter(bmi<=30  & smoker == "yes") %>%
               select(c("age", "bmi", "children", "charges")),
             main = "bmi<=30 vs charges for smoker = yes ")

pairs.panels(data %>% 
               filter(bmi<=30  & smoker == "no") %>%
               select(c("age", "bmi", "children", "charges")),
             main = "bmi<=30 vs charges for smoker = no ")
```

## Residuals Normality assumption

-   Quantile-Quantile (QQ) plots helps determine if data follows a selected distribution, here we are testing for normality; the closer the standardized residuals to the 45 degree line the closer the data is to being approximately normally distributed.

-   You give it a vector of data and R plots the data in sorted order versus quantiles from a standard Normal distribution.

-   Residuals used in QQ plots are standardized so it is easier to spot outliers (points drifting away from line)

-   Despite that 'bmi' violates residual equal variance assumption, residuals are approximately normally distributed. We are actually not looking for perfect normality rather approximate normality.

```{r}
mypar(2,2)
name = colnames(data %>% select_if(is.numeric))
for(i in name){
  qqnorm(data[[i]], main=paste0("Normal Q-Q Plot for ",i))
  qqline(data[[i]])
}
```

## Hypothesis testing
The Wilcoxon signed-rank test is a non-parametric statistical hypothesis test used to compare two related samples, matched samples, or repeated measurements on a single sample to estimate whether their population means ranks differ e.g it is a paired difference test. 

  H0: There is no difference in the distribution scores.
  HA: There is a difference in the distribution scores.

From the test below, we can observe that for smoker and divided bmi, the p values are too small (less than 5% significance level) so that we can reject the null hypothesis. We conclude that charges are significantly different between smokers and non-smokers as well as people with low bmi and people with high bmi. 

However, there is no significant difference bewteen male and female observations. (p=0.7 > 0.05)


```{r}
wilcox.test(data$charges ~ data$smoker)
wilcox.test(data$charges ~ data$sex)

bmi_category <- ifelse(data$bmi>=30, "low", "high")

wilcox.test(data$charges ~ bmi_category)
```



